Adaptative Computation:\\ An Essay in Memory of John H. Holland

### 2016-02-24 paper.tex
melanie's edits through steph

In Aug. 2015, Professor John H. Holland passed in Ann Arbor, MI, where
he had served on the University of Michigan faculty for over 50 years.
John, as he was known universally to his colleagues and students,
leaves behind a long legacy of intellectual achievements.

As a descendant of the cybernetics era, Holland was 
influenced by the work of von Neumann, Wiener, Ashby, and Turing, all
of whom viewed computation as a broad, interdisciplinary enterprise.
Thus, Holland became an early proponent of interdisciplinary
approaches to computer science and an active evangelist of what is now
called \emph{computational thinking}, reaching out enthusiastically to
psychologists, economists, physicists, linguists, philosophers, and
pretty much anyone he came in contact with.  As a result, even though
he received what was arguably one of the world's first computer
science
% SF softened the claim here after reading the comments section on the blog.
Ph.D. degrees in 1959\cite{London2003}, his contributions are sometimes better known outside of
computer science than within.

@@

Holland's aim, however, was
more general than a new class of algorithms---his goal was an
interdisciplinary theory of adaptation, one that would inform biology, 
say, as much as computer science \cite{Christiansen1998}.  The later, successful application of
genetic algorithms to real-world optimization and learning tasks was,
for Holland, just icing on the cake.  
This broader view of adaptation
has inspired many, and engendered criticism
from others, leading to spirited intellectual debates and
controversies.  Most controversial is the extent to which it can be
demonstrated, either empirically or mathematically, that the behavior
of adaptive systems is actually governed by these principles.
Regardless of one's position on this question, the ideas are compelling
and provide a set of unifying concepts for thinking about adaptation.
%as follows:
%Underlying Holland's theory of adaptation are the following core ideas:

@@

Models can be tacit and learned through evolutionary time, as in the
case of bacteria swimming up a chemical gradient, or explicit and
learned over a single lifespan, as in the case of cognitive systems
that incorporate experience into internal representations through
learning.  In Holland's view, the key activity of an adaptive agent
involves building and refining these data-driven models of the
environment.

@@

Holland's second key contribution was describing a computational
mechanism, the \emph{learning classifier
  system}~\cite{Holland1986,Holland1977}, to illustrate how a cognitive system could
iteratively build up a detailed and hierarchical model of its
environment to enhance survival.  The key learning elements of this method, the
\emph{bucket-brigade} algorithm combined with a genetic algorithm, presaged many of the ideas in modern reinforcement
learning, notably the temporal-difference methods introduced by
Sutton and Barto~\cite{SuttonAndBarto1990}.

@@

Holland's inspiration for classifier systems came from several
different disciplines, including Hebbian
learning, artificial intelligence, evolutionary biology, economics,
psychology, and control theory.
% MM says: We cite these elsewhere in the paper
%(e.g.,\cite{Bellman1961,Samuel1959}).  
Knowledge representation in the form of a population of ``if-then''
rules seemed like a good choice, not only because of its popularity in
AI at the time but also from Holland's early work on modeling Hebbian
cell assemblies: ``In Hebb's view, a cell assembly makes a simple
statement: If such and such an event occurs, then I will fire for a
while at a high rate'' \cite{Waldrop1993}.  The if-then rules, when
activated, compete to post their results on a shared ``message list,''
serving as the system's short-term memory (again inspired by Hebb's
work and AI blackboard systems of the day).  Unlike blackboard
systems, however, new rules are generated automatically in a
trial-and-error fashion, and can be selected and recombined via a
genetic algorithm.

@@

Today, we use computational models both for \emph{prediction}---by
analyzing data via statistical models---and for \emph{understanding}
how systems work---by probing the effects of hypothesized underlying
mechanisms.  This latter use of models was dear to Holland's heart.
In his view, the key to science was understanding the mechanisms that
cause a system to behave in a certain way, an aspiration that goes
well beyond data fitting methods, which typically focus only on
the aggregate behavior of a system.

For example, a purely statistical model that describes the boom and bust
pattern of the stock market
%cannot
does not address the underlying mechanisms
that lead to these cycles, through the collective actions of myriad
individual buy/sell decisions.  
%Similarly
In contrast, the genetic algorithms for which Holland is so famous
are exploratory models of \emph{mechanism}: they provide a simple
computational framework in which to explore the dynamics of Darwinian
evolution and whether the basic mechanisms of variation, differential
reproduction, and heredity are sufficient to account for the richness
of our natural world.

Holland's work focused on \emph{exploratory} models---those that explore basic principles and
mechanisms, even if they do not make specific or detailed
predictions\cite{Holland1995}.  Such models can show generically how certain behaviors
could be produced.  Holland pioneered a style of modeling that has
come to be known as ``individual-based'' or ``agent-based,'' in which
every component of a system is represented explicitly---e.g., every
trader in a stock market system or every cell in an immune system
model---and has a dynamic internal state. In such models, each agent has
its own behavior rules, which it can modify over time through learning.  In
order to capture the behavior of systems under spatial
constraints, these models are often defined over spatial structures
such as networks or simple grids.

@@

The emphasis on exploratory models to build intuitions was an
important theme of Holland's work, and he often quoted Eddington's
remark on the occasion of the first experimental test of Einstein's
theory of relativity: ``The contemplation in natural science of a wider
domain than the actual leads to a far better understanding of the
actual'' \cite{Eddington1927}.
Holland's view of modeling was by no means
typical.  As just one example of the prevailing view that models are
for predicting: 
%For example, a textbook on computational modeling offers
% the following definition: 
``Modeling is the application
of methods to analyze complex, real-world problems in order to make
predictions about what might happen with various actions''
\cite{Shiflet2014}.   This perspective rules out the
kind of exploratory modeling that interested Holland the most.

@@

Holland was interested in a broad array of adaptive systems---immune
systems, ecologies, financial markets, cities, and the brain---systems
that are \emph{complex}.  In the early 1980s, he teamed up with a small
group of scientists, primarily physicists with a sprinkling of
economists and biologists, to discuss what properties this wide swath of systems
have in common.  The discussions helped define the intellectual mission
of the Santa Fe Institute (SFI), the first institution dedicated to
developing a science of complexity, as well as the other complexity institutes that followed.
% [add in how many complexity
%  institutes there are worldwide and which others John played a hand
%  in. MM says: Do we really need that?]  
Holland brought to these discussions his lifelong study of
adaptation and a reminder that serious theories about complexity would
need to look deeper than phenomenological descriptions but also
account for the `how' and `why' of these systems.
%describing phenomena like chaos, power laws, and ONE MORE, and acoount for the

@@

ECHO \cite{Holland1999,Hraber1997} was an even more ambitious exploratory model
that Holland and collaborators developed during the 1990s. 
% ECHO's
% scope included the evolution of interacting populations of individual
% agents, which through competition and learning discover symbiotic
% triads such as the famous ant-fly-caterpillar interaction, the Wiksell
% triangle of trading relationships in economics, and the immune
% system's learned ability to distinguish `self' from `other.'
ECHO formalized Holland's idealization of complex adaptive systems
into a runnable computational system where agents evolved external
markers (called tags) and internal preferences, and then used them to
acquire resources, form higher-level aggregate structures, such as
trading relationships, symbiotic groups, trophic cascades, and
interdependent organizations.  ECHO agents sometimes discovered
mimicry and used it to deceive competitors.  In many runs, the model
developed increasingly complex structures and behaviors up to a point.
When these runs stabilized, the resulting population of agents was
found to reproduce several well-known patterns observed in nature,
most famously the rank-frequency distribution of species diversity
known as the Preston curve in ecology.  In layman's terms, ECHO
evolved populations where ``most species are rare.''

@@

Holland's later books, \emph{Emergence} \cite{Holland2000},
\emph{Signals and Boundaries} \cite{Holland2012}, and
\emph{Complexity: A Very Short Introduction} \cite{Holland2014} show how the
theories of adaptation that Holland developed during the earlier part
of his career fit into the larger landscape of complex systems
research.  In these works he especially emphasized the open-ended
nature of complex systems, where change and adaptation are
continual, systems co-evolve with each other, and ecological niches arise
and decay spontaneously depending on resource availability and competition..

@@

John Holland was unusual in his ability to absorb the essence of other
disciplines, articulate grand overarching
principles, and then back them up with computational mechanisms and
mathematics.  Unlike most researchers, Holland moved seamlessly among these
three modes of thinking, developing models that were years ahead of their time.  A
close reading of his work reveals the antecedents of many ideas prevalent in
machine learning today, e.g., reinforcement, learning in non-Markovian
environments and active learning.   His seminal genetic algorithm
spawned the field of evolutionary computation, and his insights
and wisdom helped define what are now referred to as the
\emph{sciences of complexity}.

@@

The authors thank R. Axelrod, L. Booker, R. Riolo, K. Winter, and the anonymous
reviewers for their careful reading of the manuscript and many helpful
suggestions.  SF gratefully acknowledges the partial support of NSF
(NeTS 1518878, CNS 1444500), DARPA (FA8750-15-C-0118), 
Air Force Office of Scientific Research (FA8750-15-2-0075), and the Santa Fe Institute.

>>>>

In August 2015, Professor John H. Holland passed away in Ann Arbor, MI, where
he had served on the University of Michigan faculty for over 50 years.
John, as he was known universally to his colleagues and students,
leaves behind a long legacy of intellectual achievements.

As a descendant of the cybernetics era, Holland was 
influenced by the work of von Neumann, Wiener, Ashby, and Turing, all
of whom viewed computation as a broad, interdisciplinary enterprise.
Thus, Holland became an early proponent of interdisciplinary
approaches to computer science and an active evangelist of what is now
called \emph{computational thinking}, reaching out enthusiastically to
psychologists, economists, physicists, linguists, philosophers, and
pretty much anyone he came in contact with.  As a result, even though
he received what was arguably one of the world's first computer
science
% SF softened the claim here after reading the comments section on the blog.
Ph.D. degrees in 1959 \cite{London2003}, his contributions are sometimes better known outside of
computer science than within.

@@

Holland's aim, however, was more general than a new class of algorithms.  His goal was an
interdisciplinary theory of adaptation, one that would inform biology, 
say, as much as computer science \cite{Christiansen1998}.  The later, successful application of
genetic algorithms to real-world optimization and learning tasks was,
for Holland, just icing on the cake.  
Holland's broader view of adaptation
has inspired many, and engendered criticism
from others, leading to spirited intellectual debates and
controversies.  Most controversial is the extent to which it can be
demonstrated, either empirically or mathematically, that the behavior
of adaptive systems is actually governed by Holland's proposed principles.
Regardless of one's position on this question, the ideas are compelling
and provide a set of unifying concepts for thinking about adaptation.
%as follows:
%Underlying Holland's theory of adaptation are the following core ideas:

@@

Models can be tacit and learned over evolutionary time, as in the
case of bacteria swimming up a chemical gradient, or explicit and
learned over a single lifespan, as in the case of cognitive systems
that incorporate experience into internal representations through
learning.  In Holland's view, the key activity of an adaptive agent
involves building and refining these data-driven models of the
environment.

@@

Holland's second key contribution was describing a computational
mechanism, the \emph{learning classifier
  system}~\cite{Holland1986,Holland1977}, to illustrate how a cognitive system could
iteratively build up a detailed and hierarchical model of its
environment to enhance survival.  The key learning elements of this method, the
bucket-brigade algorithm, combined with a genetic algorithm, presaged many of the ideas in modern reinforcement
learning, notably the temporal-difference methods introduced by
Sutton and Barto~\cite{SuttonAndBarto1990}.

@@

Holland's inspiration for classifier systems came from several
different disciplines, including Hebbian
learning, artificial intelligence, evolutionary biology, economics,
psychology, and control theory.
% MM says: We cite these elsewhere in the paper
%(e.g.,\cite{Bellman1961,Samuel1959}).  
Knowledge representation in the form of a population of ``if-then''
rules seemed like a good choice, not only because of its popularity in
AI at the time but also from Holland's early work on modeling Hebbian
cell assemblies: ``In Hebb's view, a cell assembly makes a simple
statement: If such and such an event occurs, then I will fire for a
while at a high rate'' \cite{Waldrop1993}.  The if-then rules, when
activated, compete to post their results on a shared ``message list,''
serving as the system's short-term memory, again inspired by Hebb's
work as well as by AI blackboard systems of the day.  Unlike blackboard
systems, however, new rules are generated automatically in a
trial-and-error fashion, and can be selected and recombined via a
genetic algorithm.

@@

Today, we use computational models both for \emph{prediction}---by
analyzing data via statistical models---and for \emph{understanding}
how systems work---by probing the effects of hypothesized underlying
mechanisms.  This latter use of models was dear to Holland's heart.
In his view, the key to science was understanding the mechanisms that
cause a system to behave in a certain way, an aspiration that goes
well beyond data-fitting methods, which typically focus only on
the aggregate behavior of a system.

For example, a purely statistical model that describes the boom and bust
pattern of the stock market
%cannot
does not address the underlying mechanisms
that lead to these cycles, through the collective actions of myriad
individual buy/sell decisions.  
%Similarly
In contrast, the genetic algorithms for which Holland is so famous
are exploratory models of \emph{mechanism}: they provide a simple
computational framework in which to explore the dynamics of Darwinian
evolution and whether the basic mechanisms of variation, differential
reproduction, and heredity are sufficient to account for the richness
of the natural world.

Holland's work focused on \emph{exploratory} models---those that explore basic principles and
mechanisms, even if they do not make specific or detailed
predictions \cite{Holland1995}.  Such models can show generically how certain behaviors
could be produced.  Holland pioneered a style of modeling that has
come to be known as ``individual-based'' or ``agent-based,'' in which
every component of a system is represented explicitly---e.g., every
trader in a stock market system or every cell in an immune system
model---and has a dynamic internal state. In such models, each agent has
its own behavior rules, which it can modify over time through learning.  In
order to capture the behavior of systems under spatial
constraints, these models are often defined over spatial structures
such as networks or simple grids.

@@

The emphasis on exploratory models to build intuitions was an
important theme of Holland's work, and he often quoted Eddington's
remark on the occasion of the first experimental test of Einstein's
theory of relativity: ``The contemplation in natural science of a wider
domain than the actual leads to a far better understanding of the
actual'' \cite{Eddington1927}.
%Holland's view of modeling was by no means
%typical.  As just one example of the prevailing view that models are
%for predicting: 
%For example, a textbook on computational modeling offers
% the following definition: 
%``Modeling is the application
%of methods to analyze complex, real-world problems in order to make
%predictions about what might happen with various actions''
%\cite{Shiflet2014}.   This perspective rules out the
%kind of exploratory modeling that interested Holland the most.

@@

Holland was interested in a broad array of adaptive systems, such as immune
systems, ecologies, financial markets, cities, and the brain---systems
that are \emph{complex}.  In the early 1980s, he teamed up with a small
group of scientists, primarily physicists, with a sprinkling of
economists and biologists, to discuss what properties this wide swath of systems
have in common.  The discussions helped define the intellectual mission
of the Santa Fe Institute (SFI), the first institution dedicated to
developing a science of complexity, as well as the other complexity institutes that followed.
% [add in how many complexity
%  institutes there are worldwide and which others John played a hand
%  in. MM says: Do we really need that?]  
Holland brought to these discussions his lifelong study of
adaptation and a reminder that serious theories about complexity would
need to look deeper than phenomenological descriptions but also
account for the `how' and `why' of these systems.
%describing phenomena like chaos, power laws, and ONE MORE, and acoount for the

@@

ECHO \cite{Holland1999,Hraber1997} was an even more ambitious exploratory model
that Holland and collaborators developed during the 1990s. 
% ECHO's
% scope included the evolution of interacting populations of individual
% agents, which through competition and learning discover symbiotic
% triads such as the famous ant-fly-caterpillar interaction, the Wiksell
% triangle of trading relationships in economics, and the immune
% system's learned ability to distinguish `self' from `other.'
ECHO formalized Holland's idealization of complex adaptive systems
into a runnable computational system where agents evolved external
markers (called tags) and internal preferences, and then used them to
acquire resources and form higher-level aggregate structures, such as
trading relationships, symbiotic groups, trophic cascades, and
interdependent organizations.  ECHO agents sometimes discovered
mimicry and used it to deceive competitors.  In many runs, the model
developed increasingly complex structures and behaviors. % up to a point.
When these runs stabilized, the resulting population of agents was
found to reproduce several well-known patterns observed in nature,
most famously the rank-frequency distribution of species diversity
known as the Preston curve in ecology.  In layman's terms, ECHO
evolved populations where ``most species are rare.''

@@

Holland's later books, \emph{Emergence} \cite{Holland2000},
\emph{Signals and Boundaries} \cite{Holland2012}, and
\emph{Complexity: A Very Short Introduction} \cite{Holland2014} show how the
theories of adaptation that Holland developed during the earlier part
of his career fit into the larger landscape of complex systems
research.  In these works he especially emphasized the open-ended
nature of complex systems, where change and adaptation are
continual, systems co-evolve with each other, and ecological niches arise
and decay spontaneously depending on resource availability and competition.

@@

John Holland was unusual in his ability to absorb the essence of other
disciplines, articulate grand overarching
principles, and then back them up with computational mechanisms and
mathematics.  Unlike most researchers, Holland moved seamlessly among these
three modes of thinking, developing models that were years ahead of their time.  A
close reading of his work reveals the antecedents of many ideas prevalent in
machine learning today, e.g., reinforcement, learning in non-Markovian
environments, and active learning.   His seminal genetic algorithm
spawned the field of evolutionary computation, and his insights
and wisdom helped define what are now referred to as the
\emph{sciences of complexity}.

@@

The authors thank R. Axelrod, L. Booker, R. Riolo, K. Winter, and the
anonymous reviewers for their careful reading of the manuscript and
many helpful suggestions.  SF gratefully acknowledges the partial
support of NSF (NeTS 1518878, CNS 1444500), DARPA (FA8750-15-C-0118),
Air Force Office of Scientific Research (FA8750-15-2-0075), and the
Santa Fe Institute.  MM gratefully acknowledges the partial support of
NSF (IIS-1423651).  Any opinions, findings, and conclusions or
recommendations expressed in this material are those of the authors
and do not necessarily reflect the views of the National Science
Foundation.
